TacticToe: Learning the Game of Tactics

### 2018-02-15 tactictoe_jv.tex
recording minor

In the next phase, we globalize identifiers of the proof scripts.
Infix operators such as \texttt{THEN} need to be processed in a special way so
that they keep their infixity status after globalization. For simplicity,
the
globalization of infix operators is omitted in Example~\ref{ex:r1}.
In this example, the three main cases that can happen during the globalization
are depicted. The first one is the globalization of identifiers declared in
modules. The global identifiers \texttt{STRIP\_TAC} and
\texttt{ASM\_REWRITE\_TAC} are prefixed by their module \texttt{Tactic}. In
this way, they will still be interpretable whether \texttt{Tactic} was open or
not. The local identifier \texttt{LIST\_INDUCT\_TAC} is replaced by its
definition which happens to contain two global identifiers.
The previous paragraph describe the globalization for all identifiers except
local theorems.  We do not replace a local
theorem by its definition because we want to avoid unfolding proof scripts into
other proof scripts.
If the theorem is stored in the database of the current theory, we
can obtain the theorem value by calling \texttt{DB.fetch}. This happens for the
three theorems in our script. The internal database of \holfour theorems is
available across \holfour developments.
Otherwise the globalization process fails and the local theorem identifier is
kept unchanged. A recorded tactic with an unchanged local theorem as argument
is only interpretable inside the current theory.

@@

\section{Case Study}\todo{comment on the case study}
Since the proofs generated by \tactictoe are meant to be part of a \holfour 
development, it interesting to compare their quality with the original human 
proofs. The quality of a proof in \holfour can be measured in terms of length, 
speed, readability and maintainability. We study these properties in three 
examples taken from our full-scale experiment.

@@

In Example~\ref{ex:cs2}, we try to prove that for any surjective function $f$ 
from $s$ to $t$, there exists an injective function $g$ from $t$ to $s$ such 
that $f o g$ is the identity function. The human proof is quite complicated. In 
contrast, \tactictoe finds a proof that first expand the definition of 
injectivity and surjectivity and then call \metis. However, the simpler script 
does take much longer to execute due to the proof search happening inside 
\metis.

\begin{example}\label{ex:cs2} \texttt{SURJ\_INJ\_INV} in theory 
\texttt{pred\_set}
\begin{lstlisting}[language=SMLSmall]
!f s t. SURJ f s t ==> ?g. INJ g t s wedge !y. y IN t ==> (f (g y) = y)
\end{lstlisting}
Human (2 milliseconds)
\begin{lstlisting}[language=SMLSmall]
REWRITE_TAC [IMAGE_SURJ] THEN
DISCH_TAC THEN Q.EXISTS_TAC `THE o LINV_OPT f s` THEN
BasicProvers.VAR_EQ_TAC THEN REPEAT STRIP_TAC THENL
  [irule INJ_COMPOSE THEN Q.EXISTS_TAC `IMAGE SOME s` THEN
     REWRITE_TAC [INJ_LINV_OPT_IMAGE] THEN REWRITE_TAC [INJ_DEF, IN_IMAGE] THEN
     REPEAT STRIP_TAC THEN REPEAT BasicProvers.VAR_EQ_TAC THEN
     FULL_SIMP_TAC std_ss [THE_DEF],
   ASM_REWRITE_TAC [LINV_OPT_def, o_THM, THE_DEF] THEN
     RULE_ASSUM_TAC (Ho_Rewrite.REWRITE_RULE
       [IN_IMAGE', GSYM SELECT_THM, BETA_THM]) THEN ASM_REWRITE_TAC []]
\end{lstlisting}
\tactictoe (50 milliseconds)
\begin{lstlisting}[language=SMLSmall]
SRW_TAC [] [SURJ_DEF, INJ_DEF] THEN METIS_TAC []
\end{lstlisting}
\end{example}

In Example~\ref{ex:cs3}, we prove a theorem about lists which is domain where 
\tactictoe excels compared to ATPs. 
Given two list $l_1$ and $l_2$ where the length of $l_1$ (noted $p$) is less 
than a natural 
number $n$, the theorem states updating the $n^{th}$ 
element of the concatenation of $l_1$ and $l_2$ is the same as
updating the $m^{th}$ element of $l_2$ where $m = n - p$. Again, 
\tactictoe's proof is 
much more readable starting with applying induction on $l$. It solves the base 
case of the induction by rewriting and distinguish cases where $n=0$ or $n>0$ 
in the inductive hypothesis. It finalizes the proof using the definition of 
$LUPDATE$. Here, \tactictoe's proof is arguably smaller, faster and easier to 
understand and maintain. Such proofs after an expert review could replace the 
their respective original human proofs in the \holfour repository.

\begin{example}\label{ex:cs3} \texttt{LUPDATE\_APPEND2} in theory 
\texttt{rich\_list}
\begin{lstlisting}[language=SMLSmall]
!l1 l2 n x. LENGTH l1 <= n ==>
  (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n - LENGTH l1) l2))
\end{lstlisting}
Human (63 milliseconds)
\begin{lstlisting}[language=SMLSmall]
  rw[] THEN simp[LIST_EQ_REWRITE] THEN Q.X_GEN_TAC `z` THEN
  simp[EL_LUPDATE] THEN rw[] THEN simp[EL_APPEND2,EL_LUPDATE] THEN
  fs[] THEN Cases_on `z < LENGTH l1` THEN
  fs[] THEN simp[EL_APPEND1,EL_APPEND2,EL_LUPDATE]
\end{lstlisting}
\tactictoe (17 milliseconds)
\begin{lstlisting}[language=SMLSmall]
Induct_on `l1` THENL [SRW_TAC [] [],
  Cases_on `n` THENL [SRW_TAC [] [],
    FULL_SIMP_TAC (srw_ss ()) [] THEN METIS_TAC [LUPDATE_def]]]
\end{lstlisting}
\end{example}

>>>>

In the next phase, we globalize identifiers of the proof scripts.
Infix operators such as \texttt{THEN} need to be processed in a special way so
that they keep their infixity status after globalization. For simplicity,
the
globalization of infix operators is omitted in Example~\ref{ex:r1}.
In this example, the three main cases that can happen during the globalization
are depicted. The first one is the globalization of identifiers declared in
modules. The global identifiers \texttt{STRIP\_TAC} and
\texttt{ASM\_REWRITE\_TAC} are prefixed by their module \texttt{Tactic}. In
this way, they will still be interpretable whether \texttt{Tactic} was open or
not. The local identifier \texttt{LIST\_INDUCT\_TAC} is replaced by its
definition which happens to contain two global identifiers.
The previous paragraph describe the globalization for all identifiers except
local theorems.  We do not replace a local
theorem by its definition because we want to avoid unfolding proof scripts into
other proof scripts.
If the theorem is stored in the \holfour database available across \holfour 
developments, we can obtain the theorem value by calling \texttt{DB.fetch}.
Otherwise the globalization process fails and the local theorem identifier is
kept unchanged. A recorded tactic with an unchanged local theorem as argument
is only interpretable inside the current theory.

@@

\section{Case Study}\todo{comment on the case study}
Since the proofs generated by \tactictoe are meant to be part of a \holfour 
development, it interesting to compare their quality with the original human 
proofs. The quality of a proof in \holfour can be measured in terms of length, 
, readability, maintainability and verification speed. We study these 
properties 
in three examples taken from our full-scale experiment. The time needed by 
\holfour to check a proof is shown in parentheses.

@@

In Example~\ref{ex:cs2}, we try to prove that for any surjective function $f$ 
from $s$ to $t$, there exists an injective function $g$ from $t$ to $s$ such 
that $f\ o\ g$ is the identity function. The human proof is quite complicated. 
In 
contrast, \tactictoe finds a much smaller proof that expands the 
definition of injectivity and surjectivity and calls \metis. However, the 
\tactictoe's script does take much longer to check due to the proof search 
happening inside 
\metis.

\begin{example}\label{ex:cs2} \texttt{SURJ\_INJ\_INV} in theory 
\texttt{pred\_set}
\begin{lstlisting}[language=SMLSmall]
!f s t. SURJ f s t ==> ?g. INJ g t s wedge !y. y IN t ==> (f (g y) = y)
\end{lstlisting}
Human proof (2 milliseconds)
\begin{lstlisting}[language=SMLSmall]
REWRITE_TAC [IMAGE_SURJ] THEN
DISCH_TAC THEN Q.EXISTS_TAC `THE o LINV_OPT f s` THEN
BasicProvers.VAR_EQ_TAC THEN REPEAT STRIP_TAC THENL
  [irule INJ_COMPOSE THEN Q.EXISTS_TAC `IMAGE SOME s` THEN
     REWRITE_TAC [INJ_LINV_OPT_IMAGE] THEN REWRITE_TAC [INJ_DEF, IN_IMAGE] THEN
     REPEAT STRIP_TAC THEN REPEAT BasicProvers.VAR_EQ_TAC THEN
     FULL_SIMP_TAC std_ss [THE_DEF],
   ASM_REWRITE_TAC [LINV_OPT_def, o_THM, THE_DEF] THEN
     RULE_ASSUM_TAC (Ho_Rewrite.REWRITE_RULE
       [IN_IMAGE', GSYM SELECT_THM, BETA_THM]) THEN ASM_REWRITE_TAC []]
\end{lstlisting}
\tactictoe proof (50 milliseconds)
\begin{lstlisting}[language=SMLSmall]
SRW_TAC [] [SURJ_DEF, INJ_DEF] THEN METIS_TAC []
\end{lstlisting}
\end{example}

In Example~\ref{ex:cs3}, we prove a theorem about lists which is domain where 
\tactictoe excels compared to ATPs. 
Given two list $l_1$ and $l_2$ where the length of $l_1$ (noted $p$) is less 
than a natural 
number $n$, the theorem states updating the $n^{th}$ 
element of the concatenation of $l_1$ and $l_2$ is the same as
updating the $m^{th}$ element of $l_2$ where $m = n - p$. Again, 
\tactictoe's proof is 
much more readable starting with applying induction on $l$. It solves the base 
case of the induction by rewriting and distinguish cases where $n=0$ or $n>0$ 
in the inductive hypothesis. It finalizes the proof with a short call to \metis 
using the definition of 
$LUPDATE$. Here, \tactictoe's proof is arguably smaller, faster and easier to 
understand and maintain. Such proofs after an expert review could replace the 
their respective original human proofs in the \holfour repository.

\begin{example}\label{ex:cs3} \texttt{LUPDATE\_APPEND2} in theory 
\texttt{rich\_list}
\begin{lstlisting}[language=SMLSmall]
!l1 l2 n x. LENGTH l1 <= n ==>
  (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n - LENGTH l1) l2))
\end{lstlisting}
Human proof (63 milliseconds)
\begin{lstlisting}[language=SMLSmall]
  rw[] THEN simp[LIST_EQ_REWRITE] THEN Q.X_GEN_TAC `z` THEN
  simp[EL_LUPDATE] THEN rw[] THEN simp[EL_APPEND2,EL_LUPDATE] THEN
  fs[] THEN Cases_on `z < LENGTH l1` THEN
  fs[] THEN simp[EL_APPEND1,EL_APPEND2,EL_LUPDATE]
\end{lstlisting}
\tactictoe proof (17 milliseconds)
\begin{lstlisting}[language=SMLSmall]
Induct_on `l1` THENL [SRW_TAC [] [],
  Cases_on `n` THENL [SRW_TAC [] [],
    FULL_SIMP_TAC (srw_ss ()) [] THEN METIS_TAC [LUPDATE_def]]]
\end{lstlisting}
\end{example}
